---
title: "Cluster tendency using unsupervised random forest"
author: "Yi Chen"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_depth: 3
        toc_float:
            collapsed: true
        number_sections: true
        code_download: true
        df_print: kable
        code_folding: hide
        mode: selfcontained
---
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>
<style type="text/css">
body{ font-family: serif; font-size: 11pt; }
h1 { font-size: 24pt; }
h2 { font-size: 20pt; }
h3 { font-size: 16pt; }
summary { cursor: pointer; }
p { max-width: 600px; }
caption, .caption { color: #666666; }
details {padding-top:15px; padding-bottom:15px;}
embed { width: 750px; height: 600px; }
.toc-content{ max-width: 750px; }
</style>
```{r setup, echo=F, warning=F, setup=T}
knitr::opts_knit$set(upload.fun = knitr::image_uri)
knitr::opts_chunk$set(fig.align="center", results="hold", cache.path="cache/", dev="png",
                      fig.width=6.5, fig.height=6.5)
R.utils::setOption("digits", 4)
stderrp = function(...) { 
    dots = list(...)
    if(any(sapply(dots, class) == 'character')) {
        write(paste(...), stderr()) 
    } else {
        write(paste0(capture.output(...)), stderr()) 
    }
}
ic = stderrp
knitr::knit_hooks$set(timeit = local({
    now <- NULL
    function(before, options) {
        if(before) {
            now <<- Sys.time()
        } else {
            res <- difftime(Sys.time(), now)
            now <- NULL
            stderrp(paste('\n',
                stringr::str_pad(options$label, 20, side="right"), 
                stringr::str_pad(round(as.numeric(res), digits=2), 9),
                units(res)))
        }
    }}),
    summary = {
        function(before, options) {
            if(before) {
                return(asis_output(paste0("<details><summary>", 
                                         gsub('_', ' ', options$label), "</summary>\n")))
            } else {
                return(asis_output("</details>\n"))
            }
        }
    }
)
knitr::opts_chunk$set(timeit=T)
is.html = knitr::is_html_output()
```
```{r loadLibraries, echo=F, message=F, warning=F, loadLibraries=T}
library(xfun)
library(knitr)
library(reactable)
library(tidyverse)
library(patchwork)
library(glmnet)
```

# Introduction

Previously, we showed that the data followed a Beta distribution when the power of the Hopkins
statistic was reduced to $1.5\log{d}\log{m}$. There seemd to be some bias in this empirical result
based on visual examination. Here, I examine it further.

```{r loadData}
raw_dat = readRDS("cache/beta_test6_d253efd2a1fdfc81099175ecf66990e3.rds")
```

# Varying the power

```{r test_power_func}
dvals = c(5, 10, 20, 50, 100, 200)
mvals = c(5, 10, 20, 50, 100, 200)
p = ((1:200)/200 - 1/400)

comp_dists = function(x, x_target) {
    sum(abs(x - x_target))
}
hopcalc = function(obj, d=1) {
    y_i = obj$y_i^d
    x_i = obj$x_i^d
    sum(y_i) / (sum(x_i) + sum(y_i))
}
test_power = function(i, j, dat, print.out=T) {
    d = dvals[i]
    m = mvals[j]
    test_powers = log(d)*log(m)*seq(1, 2, length.out=100)
    target_dist = qbeta(p, shape1=m, shape2=m)
    tmp = sapply(test_powers, function(dt) {
        tmp = sapply(1:200, function(s) hopcalc(dat[[i]][[j]][[s]], dt))
        comp_dists(sort(tmp), target_dist)
    })
    if(print.out) {
        return(rbind(sum_diff=tmp, power=test_powers, true_d=d))
    }
    data.frame(d=d, m=m, best_power=test_powers[which.min(tmp)])
}
```
```{r vary_power, summary=T}
tmp = cache_rds({
    bind_rows(lapply(1:6, function(i) 
                     bind_rows(lapply(1:6, function(j) test_power(i, j, raw_dat, print.out=F)))))
}, rerun=F, clean=T)
tmp
vary_power = tmp
```

```{r cbrt_func}
cbrt = function(x) x^(1/3)
```
```{r vary_power_plot5}
tmp2 = bind_rows(
tmp %>% filter(m == 5) %>% mutate(group="m=5", varied=cbrt(d)),
tmp %>% filter(d == 5) %>% mutate(group="d=5", varied=cbrt(m))
)
vp5 = ggplot(tmp2) +
    aes(x=varied, y=best_power, color=group, group=group) +
    geom_line() +
    geom_point() +
    xlab("cube root(m or d)") +
    scale_color_brewer("", palette="Dark2") +
    geom_line(stat="smooth", method="lm", formula = y ~ x, se = F, linetype="dashed", alpha=0.5) +
    theme_minimal()
```
```{r vary_power_plot10}
tmp2 = bind_rows(
tmp %>% filter(m == 10) %>% mutate(group="m=10", varied=cbrt(d)),
tmp %>% filter(d == 10) %>% mutate(group="d=10", varied=cbrt(m))
)
vp10 = ggplot(tmp2) +
    aes(x=varied, y=best_power, color=group, group=group) +
    geom_line() +
    geom_point() +
    xlab("cube root(m or d)") +
    scale_color_brewer("", palette="Dark2") +
    geom_line(stat="smooth", method="lm", formula = y ~ x, se = F, linetype="dashed", alpha=0.5) +
    theme_minimal()
```
```{r vary_power_plot20}
tmp2 = bind_rows(
tmp %>% filter(m == 20) %>% mutate(group="m=20", varied=cbrt(d)),
tmp %>% filter(d == 20) %>% mutate(group="d=20", varied=cbrt(m))
)
vp20 = ggplot(tmp2) +
    aes(x=varied, y=best_power, color=group, group=group) +
    geom_line() +
    geom_point() +
    xlab("cube root(m or d)") +
    scale_color_brewer("", palette="Dark2") +
    geom_line(stat="smooth", method="lm", formula = y ~ x, se = F, linetype="dashed", alpha=0.5) +
    theme_minimal()
```
```{r vary_power_plot50}
tmp2 = bind_rows(
tmp %>% filter(m == 50) %>% mutate(group="m=50", varied=cbrt(d)),
tmp %>% filter(d == 50) %>% mutate(group="d=50", varied=cbrt(m))
)
vp50 = ggplot(tmp2) +
    aes(x=varied, y=best_power, color=group, group=group) +
    geom_line() +
    geom_point() +
    xlab("cube root(m or d)") +
    scale_color_brewer("", palette="Dark2") +
    geom_line(stat="smooth", method="lm", formula = y ~ x, se = F, linetype="dashed", alpha=0.5) +
    theme_minimal()
```
```{r vary_power_plot100}
tmp2 = bind_rows(
tmp %>% filter(m == 100) %>% mutate(group="m=100", varied=cbrt(d)),
tmp %>% filter(d == 100) %>% mutate(group="d=100", varied=cbrt(m))
)
vp100 = ggplot(tmp2) +
    aes(x=varied, y=best_power, color=group, group=group) +
    geom_line() +
    geom_point() +
    xlab("cube root(m or d)") +
    scale_color_brewer("", palette="Dark2") +
    geom_line(stat="smooth", method="lm", formula = y ~ x, se = F, linetype="dashed", alpha=0.5) +
    theme_minimal()
```
```{r show_vary_plots, fig.height=7}
vp5 + vp10 + vp20 + vp50 + vp100 + plot_layout(ncol=2)
```

<mark>Observations:</mark>

- When $d$ is constant, the growth w/r/t $m$ is faster than when $m$ is constant.
- It seems that the best power appears to be a function of $m^{\frac{1}{3}}$, where the slope may
  be a function of $d$.

## Finding the slope when m is constant

```{r dconstant_slope}
dconstant = bind_rows(lapply(dvals, function(dval) {
    tmp = tmp %>% filter(d == dval) %>% mutate(cbrt_m = cbrt(m))
    mod = lm(best_power ~ cbrt_m, data=tmp)
    mod2 = lm(best_power ~ cbrt_m + 0, data=tmp)
    data.frame(d=dval, slope=coef(mod)[2], intercept=coef(mod)[1], slope_0=coef(mod2)[1], 
               r2=summary(mod)$r.squared, r2_0=summary(mod2)$r.squared)
}))
rownames(dconstant) = dvals
dconstant
```
```{r d_slope_plot}
tmp = dconstant %>%
    mutate(log_d=log(d), sqrt_d=sqrt(d), cbrt_d=cbrt(d), 
           mtry=floor(sqrt(d)), log_mtry=log(floor(sqrt(d))))
d_transforms = c("d", "log_d", 'sqrt_d', 'cbrt_d', 'mtry', 'log_mtry')
plots = lapply(d_transforms, function(x) {
    ggplot(tmp) +
        aes(x=.data[[x]], y=slope_0) +
        geom_point() +
        xlab(x) +
        geom_line(stat="smooth", method="lm", formula=y ~ x, se=F, linetype="dashed", alpha=0.5) +
        theme_minimal()
})
wrap_plots(plots, ncol=2)
```
```{r log_d_slope_linear_model, summary=T}
mod = lm(slope_0 ~ log_d, data=tmp)
summary(mod)
```
```{r log_mtry_slope_linear_model, summary=T}
mod = lm(slope_0 ~ log_mtry, data=tmp)
summary(mod)
```

```{r}
vary_power = vary_power %>%
    mutate(mtry=floor(sqrt(d)), log_mtry=log(mtry), pred=(3.642*log_mtry-1)*cbrt(m))
p1 = ggplot(vary_power) +
    aes(x=best_power, y=pred, color=factor(m)) +
    scale_color_brewer("m", palette="Dark2") +
    geom_point() +
    geom_line() +
    geom_abline(slope=1, intercept=0, linetype="dashed", alpha=0.5) +
    theme_minimal()
p2 = ggplot(vary_power) +
    aes(x=best_power, y=pred, color=factor(d)) +
    scale_color_brewer("d", palette="Set2") +
    geom_point() +
    geom_line() +
    geom_abline(slope=1, intercept=0, linetype="dashed", alpha=0.5) +
    theme_minimal()
p1 + p2
```
