---
title: "Comparing variable importance orderings across k values"
author: "Yi Chen"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_depth: 3
        toc_float:
            collapsed: true
        number_sections: true
        code_download: true
        df_print: kable
        code_folding: hide
        mode: selfcontained
---
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>
<style type="text/css">
body{ font-family: serif; font-size: 11pt; }
h1 { font-size: 24pt; }
h2 { font-size: 20pt; }
h3 { font-size: 16pt; }
summary { cursor: pointer; }
p { max-width: 600px; }
caption, .caption { color: #666666; }
details {padding-top:15px; padding-bottom:15px;}
embed { width: 750px; height: 600px; }
.toc-content{ max-width: 750px; }
</style>
```{r setup, echo=F, warning=F, setup=T}
knitr::opts_knit$set(upload.fun = knitr::image_uri)
knitr::opts_chunk$set(fig.align="center", results="hold", cache.path="cache/", dev="png",
                      fig.width=6.5, fig.height=6.5)
R.utils::setOption("digits", 4)
stderrp = function(...) { 
    dots = list(...)
    if(any(sapply(dots, class) == 'character')) {
        write(paste(...), stderr()) 
    } else {
        write(paste0(capture.output(...)), stderr()) 
    }
}
ic = stderrp
knitr::knit_hooks$set(timeit = local({
    now <- NULL
    function(before, options) {
        if(before) {
            now <<- Sys.time()
        } else {
            res <- difftime(Sys.time(), now)
            now <- NULL
            stderrp(paste('\n',
                stringr::str_pad(options$label, 20, side="right"), 
                stringr::str_pad(round(as.numeric(res), digits=2), 9),
                units(res)))
        }
    }}),
    summary = {
        function(before, options) {
            if(before) {
                return(asis_output(paste0("<details><summary>", 
                                         gsub('_', ' ', options$label), "</summary>\n")))
            } else {
                return(asis_output("</details>\n"))
            }
        }
    }
)
knitr::opts_chunk$set(timeit=T)
is.html = knitr::is_html_output()
```
```{r loadLibraries, echo=F, message=F, warning=F, loadLibraries=T}
library(xfun)
library(yfun)
library(knitr)
library(reactable)
library(tidyverse)
library(patchwork)
library(parallel)
source('../R/RCE_algorithm.R')
```

# Introduction

[Link to RCE paper](https://link.springer.com/article/10.1007/s10994-013-5337-8)

This document investigates how different $k$ values changes the variable importance orderings
produced by RCE. This is quite important as we likely do not know the expected number of clusters
*a priori*.

# Parkinsons data set

True $k = 2$

```{r parkinsons_ktest}
data <- read.table("../example_data/Bases/parkinson.txt", header=F, sep = "\t")
k = ncol(data)
V <- data[, k] # True clustering
true_nclusters <- max(V) # Number of clusters
data <- data[, 1:(k - 1)]

parkinson_ktest = cache_rds({
    lapply(2:10, function(k) {
        mclapply(1:200, function(j) run_RCE(data, nclusters=k, nbiter=200, vfun=ic)$imp, mc.cores=4)
    })
}, rerun=F, clean=T)
names(parkinson_ktest) = paste('k =', 2:10)
```

```{r pairwise_cor_parkinsons}
tmp = sapply(parkinson_ktest, function(x) {
    apply(do.call('cbind', x), 1, mean)
})
tmp = cor(tmp, method = 'spearman')
tmp[lower.tri(tmp, diag=T)] = NA
tmp2 = data.frame(tmp, check.names=F) %>%
    tibble::rownames_to_column('k') %>%
    pivot_longer(!k) %>%
    filter(!is.na(value)) %>%
    mutate(name = factor(name, levels = names(parkinson_ktest))) %>%
    mutate(k = factor(k, levels = names(parkinson_ktest)))
p1 = ggplot(tmp2) +
    geom_tile(aes(x = k, y = name, fill = value)) +
    scale_fill_gradient(name="", low="white", high='steelblue', limits=c(0, 1)) +
    xlab("") +
    ylab("") +
    ggtitle("Spearman Correlation") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=60, vjust=1, hjust=1))
min_park_spear = min(tmp2$value)
```
```{r pairwise_cor_parkinsons2, fig.height=4, fig.width=8}
tmp = sapply(parkinson_ktest, function(x) {
    apply(do.call('cbind', x), 1, mean)
})
tmp = cor(tmp, method = 'pearson')
tmp[lower.tri(tmp, diag=T)] = NA
tmp2 = data.frame(tmp, check.names=F) %>%
    tibble::rownames_to_column('k') %>%
    pivot_longer(!k) %>%
    filter(!is.na(value)) %>%
    mutate(name = factor(name, levels = names(parkinson_ktest))) %>%
    mutate(k = factor(k, levels = names(parkinson_ktest)))
p2 = ggplot(tmp2) +
    geom_tile(aes(x = k, y = name, fill = value)) +
    scale_fill_gradient(name="", low="white", high='green4', limits=c(0, 1)) +
    xlab("") +
    ylab("") +
    ggtitle("Pearson Correlation") +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle=60, vjust=1, hjust=1))
p1 + p2
min_park = min(tmp2$value)
```


# Breast Tissue

True $k = 6$; however, one source claims that 3 classes are not distinguishable using this data
set. i.e., $k = 4$

```{r bt_ktest}
data <- read.table("../example_data/Bases/Breast_Tissu.txt", header=F, sep = "\t")
k = ncol(data)
V <- data[, k] # True clustering
true_nclusters <- max(V) # Number of clusters
data <- data[, 1:(k - 1)]

parkinson_ktest = cache_rds({
    lapply(2:10, function(k) {
        mclapply(1:200, function(j) run_RCE(data, nclusters=k, nbiter=200, vfun=ic)$imp, mc.cores=4)
    })
}, rerun=F, clean=T)
names(parkinson_ktest) = paste('k =', 2:10)
```

```{r pairwise_cor_bt}
tmp = sapply(parkinson_ktest, function(x) {
    apply(do.call('cbind', x), 1, mean)
})
tmp = cor(tmp, method = 'spearman')
tmp[lower.tri(tmp, diag=T)] = NA
tmp2 = data.frame(tmp, check.names=F) %>%
    tibble::rownames_to_column('k') %>%
    pivot_longer(!k) %>%
    filter(!is.na(value)) %>%
    mutate(name = factor(name, levels = names(parkinson_ktest))) %>%
    mutate(k = factor(k, levels = names(parkinson_ktest)))
p1 = ggplot(tmp2) +
    geom_tile(aes(x = k, y = name, fill = value)) +
    scale_fill_gradient(name="", low="white", high='steelblue', limits=c(0, 1)) +
    xlab("") +
    ylab("") +
    ggtitle("Spearman Correlation") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=60, vjust=1, hjust=1))
min_bt_spear = min(tmp2$value)
```
```{r pairwise_cor_bt2, fig.height=4, fig.width=8}
tmp = sapply(parkinson_ktest, function(x) {
    apply(do.call('cbind', x), 1, mean)
})
tmp = cor(tmp, method = 'pearson')
tmp[lower.tri(tmp, diag=T)] = NA
tmp2 = data.frame(tmp, check.names=F) %>%
    tibble::rownames_to_column('k') %>%
    pivot_longer(!k) %>%
    filter(!is.na(value)) %>%
    mutate(name = factor(name, levels = names(parkinson_ktest))) %>%
    mutate(k = factor(k, levels = names(parkinson_ktest)))
p2 = ggplot(tmp2) +
    geom_tile(aes(x = k, y = name, fill = value)) +
    scale_fill_gradient(name="", low="white", high='green4', limits=c(0, 1)) +
    xlab("") +
    ylab("") +
    ggtitle("Pearson Correlation") +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle=60, vjust=1, hjust=1))
p1 + p2
min_bt = min(tmp2$value)
```

# Conclusions

- Generally, a range of $k$ values around the true number of clusters ($k=2$ for Parkinsons; $k=6$
  for Breast Tissue) result in similar variable importance rankings.
- Minimum Spearman correlation is `r min_park_spear` for Parkinsons and `r min_bt_spear` for Breast
  Tissue.
- Minimum Pearson correlation is `r min_park` for Parkinsons and `r min_bt` for Breast Tissue.
- The rank order of variable importance appears to be quite stable across different $k$ values,
  which means that a forward/recursive search would likely be robust to the choice of $k$.
- However, the lower Pearson correlation may indicate that using the scree test as a hard cutoff
  for variables would potentially yield different sets across $k$ values.

