---
title: "Dermatology tests"
author: "Yi Chen"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: paper
        highlight: tango
        toc: true
        toc_depth: 3
        toc_float:
            collapsed: true
        number_sections: true
        code_download: true
        df_print: kable
        code_folding: hide
        mode: selfcontained
---
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>
<style type="text/css">
body{ font-family: serif; font-size: 11pt; }
h1 { font-size: 24pt; }
h2 { font-size: 20pt; }
h3 { font-size: 16pt; }
summary { cursor: pointer; }
p { max-width: 600px; }
caption, .caption { color: #666666; }
details {padding-top:15px; padding-bottom:15px;}
embed { width: 750px; height: 600px; }
.toc-content{ max-width: 750px; }
</style>
```{r setup, echo=F, warning=F, setup=T}
knitr::opts_knit$set(upload.fun = knitr::image_uri)
knitr::opts_chunk$set(fig.align="center", results="hold", cache.path="cache/", dev="png",
                      fig.width=6.5, fig.height=6.5)
R.utils::setOption("digits", 4)
stderrp = function(...) { 
    dots = list(...)
    if(any(sapply(dots, class) == 'character')) {
        write(paste(...), stderr()) 
    } else {
        write(paste0(capture.output(...)), stderr()) 
    }
}
ic = stderrp
knitr::knit_hooks$set(timeit = local({
    now <- NULL
    function(before, options) {
        if(before) {
            now <<- Sys.time()
        } else {
            res <- difftime(Sys.time(), now)
            now <- NULL
            stderrp(paste('\n',
                stringr::str_pad(options$label, 20, side="right"), 
                stringr::str_pad(round(as.numeric(res), digits=2), 9),
                units(res)))
        }
    }}),
    summary = {
        function(before, options) {
            if(before) {
                return(asis_output(paste0("<details><summary>", 
                                         gsub('_', ' ', options$label), "</summary>\n")))
            } else {
                return(asis_output("</details>\n"))
            }
        }
    }
)
knitr::opts_chunk$set(timeit=T)
is.html = knitr::is_html_output()
```



```{r loadLibraries, echo=F, message=F, warning=F, loadLibraries=T}
library(xfun)
library(knitr)
library(reactable)
library(tidyverse)
library(patchwork)
library(randomForest)
library(reshape2)
library(ggplot2)
library(pheatmap)
library(aricode)
library(mclust)
library(ess)
```

# Code

```{r clustering_module, class.source='fold-show', summary=T}
#' Clustering module
#'
#' Takes as input a distance matrix and runs clustering algorithms with a given set of parameters.
#'
#' @param x distance matrix object of type `dist` with dimensions n x n
#' @param algo clustering algorithm
#' @param k_method method for determining the number of clusters (ignored for OPTICS)
#' @param ... additional arguments depending on the choice of `algo`
#'  - For 'OPTICS': 
#'     \itemize{
#'       \item \code{minPts}: Minimum number of points in a neighborhood for a point to be considered a core point.
#'       \item \code{xi}: Minimum reachability distance for clustering.
#'     }
#'   - For 'kmeans': 
#'     \itemize{
#'       \item \code{k}: Number of clusters or initial cluster centers.
#'     }
#'   - For 'hierarchical': 
#'     \itemize{
#'       \item \code{agg_method}: Linkage method ('complete', 'single', 'average', etc.) for hierarchical clustering.
#'       \item \code{k}: Number of clusters to be formed by cutting the dendrogram.
#'     }
#' @return vector of length n with cluster memberships
#' @import dbscan
#' @import cluster
#' @export
#' 
cluster_module <- function(x, 
                           algo=c('OPTICS', 'kmeans', 'pam', 'fanny', 'hierarchical', 'agnes', 'diana'),
                           ...) {
  algo = match.arg(algo)
  params <- list(...)
  cvec <- vector()
  switch(algo,
         'OPTICS' = {
           # Ordering Points to Identify the Clustering Structure (OPTICS)
           reach <- dbscan::optics(x, minPts = params$minPts)
           clusters <- dbscan::extractXi(reach, xi= params$xi, minimum=T)
           cvec <- clusters$cluster
         },
         'kmeans' = {
           # K-Means Clustering (Partitioning Clustering)
           clusters <- stats::kmeans(x, centers = params$k)
           cvec <- clusters$cluster
         },
         'pam' = {
           # Partitioning Around Medoids (Partitioning Clustering)
           clusters <- cluster::pam(x, k = params$k)
           cvec <- clusters$clustering
         },
         'fanny' = {
           # Fuzzy Analysis Clustering (Partitioning Clustering)
           clusters <- cluster::fanny(x, k = params$k, maxit=1000)
           cvec <- clusters$clustering
         },
         'hierarchical' = {
           # Hierarchical Clustering
           switch(params$agg_method,
                  'average' = {
                    clusters <- stats::hclust(x, method = "average")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'single' = {
                    clusters <- stats::hclust(x, method = "single")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'complete' = {
                    clusters <- stats::hclust(x, method = "complete")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'ward' = {
                    clusters <- stats::hclust(x, method = "ward.D")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'mcquitty' = {
                    clusters <- stats::hclust(x, method = "mcquitty")
                    cvec <- stats::cutree(clusters, k = params$k)
                  })
         },
         'agnes' = {
           # Agglomerative Nesting (Hierarchical Clustering)
           switch(params$agg_method,
                  'average' = {
                    clusters <- cluster::agnes(x, method = "average")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'single' = {
                    clusters <- cluster::agnes(x, method = "single")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'complete' = {
                    clusters <- cluster::agnes(x, method = "complete")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'ward' = {
                    clusters <- cluster::agnes(x, method = "ward")
                    cvec <- stats::cutree(clusters, k = params$k)
                  },
                  'mcquitty' = {
                    clusters <- cluster::agnes(x, method = "weighted")
                    cvec <- stats::cutree(clusters, k = params$k)
                  })
         },
         'diana' = {
           # Divisive Analysis Clustering (Hierarchical Clustering)
           clusters <- cluster::diana(x)
           cvec <- stats::cutree(clusters, k = params$k)
         },
         {
           stop("The argument k_method is required. ", "Possible values are OPTICS, kmeans, pam, fanny, hierarchical, agnes, diana")
         })
  return(cvec)
}

```

```{r evaluation_module, class.source='fold-show', summary=T}
#' Evaluation module
#'
#' Takes as input the cluster memberships produced by a single run of `cluster_module`, and the
#' parameters used by `cluster_module`. Returns a list of evaluation metrics for that run of
#' `cluster_module`.
#'
#' @param cvec numeric vector of length n with cluster memberships
#' @param parameter list of parameters used by `cluster_module`
#' @param supervised boolean indicating whether or not to use supervised learning as a metric of
#'   cluster quality
#' @return numeric vector of evaluation metrics including
#'   - Cohesiveness: Measures how tightly grouped points are within clusters.
#'   - Separation: Measures how distinct clusters are from each other.
#'   - Cluster Sizes: Number of data points in each cluster.
#'   - Other relevant clustering metrics depending on the implementation.
#' @import fpc
#' @import clusterSim
#' @export
#' 
evaluation_module <- function(x,
                              cvec, 
                              parameter, 
                              supervised=F) {
  if(is.null(cvec) | all(cvec == 0)) { 
      return(list(nbclust=NA,
                  cohesiveness=NA,
                  distinctiveness=NA,
                  n_between=NA,
                  n_within=NA,
                  avg_silwidth=NA,
                  max_diameter=NA,
                  min_separation=NA,
                  pearson_gamma=NA,
                  dunn_index=NA,
                  dunn2_index=NA,
                  entropy=NA,
                  wb_ratio=NA,
                  ch_index=NA,
                  widest_gap=NA,
                  sep_index=NA,
                  index_C=NA,
                  index_G2=NA,
                  index_G3=NA,
                  index_S=NA
      )) 
  }
  # calculate clustering statistics
  stats <- fpc::cluster.stats(x, cvec)
  
  # prepare output with evaluation metrics
  out = list(nbclust = stats$cluster.number,
             cohesiveness = stats$average.within, 
             # Measures how closely points are grouped within the same cluster.
             distinctiveness = stats$average.between, 
             # Measures how well clusters are separated from each other.
             n_between = stats$n.between,
             # Indicates the number of between-cluster distance measurements.
             n_within = stats$n.within,
             # Indicates the number of within-cluster distance measurements.
             avg_silwidth = stats$avg.silwidth,
             #  Provides the average silhouette width, which measures the quality of the clusters.
             max_diameter = stats$max.diameter,
             # Measures the greatest distance within two points in a cluster
             min_separation = stats$min.separation,
             # Measures the minimum separation between clusters.
             # gamma_coef = stats$g2,
             # Represents the gamma coefficient, used to evaluate clustering quality.
             # g3_coeff = stats$g3,
             # Represents the g3 coefficient, used to evaluate clustering quality.
             pearson_gamma = stats$pearsongamma,
             # Represents the pearson gamma coefficient which assess clustering quality. 
             dunn_index = stats$dunn,
             # Evaluates the ratio of minimum inter-cluster distance to maximum intra-cluster distance.
             dunn2_index = stats$dunn2,
             # Measures the modified Dunn index.
             entropy = stats$entropy,
             # Measures the entropy (distance and homogenity) of the clusters
             wb_ratio = stats$wb.ratio,
             # Represents the within-between cluster ratio, used for clustering evaluation.
             ch_index = stats$ch,
             # Provides the Calinski-Harabasz index, used to evaluate the dispersion of clusters.
             widest_gap = stats$widestgap,
             # Measures the widest gap between clusters.
             sep_index = stats$sindex,
             # Represents the separation index, used for clustering evaluation.
             index_C = clusterSim::index.C(x, cvec),
             # Calculates Hubert & Levin C index - internal cluster quality index
             index_G2 = clusterSim::index.G2(x, cvec),
             # Calculates G2 internal cluster quality index
             index_G3 = clusterSim::index.G3(x, cvec),
             # Calculates G3 internal cluster quality index
             index_S = clusterSim::index.S(x, cvec)
             # Calculates Rousseeuw’s Silhouette internal cluster quality index
  )
  
  #assign class to output list
  class(out) = 'evaluation_module'
  return(out)
}
```

```{r run_all_function, class.source='fold-show', summary=T}
run_all = function(data, 
                   labels,
                   k=seq(2, 10), 
                   agg = c('average','single','complete','ward','mcquitty'),
                   xi=c(0.01, 0.05, 0.1, 0.2, 0.5), 
                   minPts=4:10) {
  data.dist = dist(data, method="euclidean")
  #data.rf <- randomForest::randomForest(x=data, proximity=TRUE, keep.forest=FALSE, ntree=2000)
  #data.dist = 1 - data.rf$proximity
  

    algos = eval(formals(args(cluster_module))$algo)
    df1 = lapply(k, function(kval) {
        lapply(algos, function(algo) {
            if(algo == "OPTICS" | algo == "hierarchical" | algo == "agnes") { return(NULL) }
            cvec = cluster_module(data.dist, algo=algo, k=kval, agg_method="complete")
            names(cvec) = paste0("sample",1:length(cvec))
            c(param.algo=algo, param.k=kval, evaluation_module(data.dist, cvec), 
             NMI=aricode::NMI(unlist(cvec), labels),
             ARI=mclust::adjustedRandIndex(unlist(cvec), labels),
             cvec)
        })
    })
    df1 = bind_rows(unlist(df1, recursive=F))
    
    # optics_params = expand.grid(xi=xi, minPts=minPts)
    # df2 = apply(optics_params, 1, function(row) {
    #   algo = "OPTICS"
    #   xival = row[["xi"]]
    #   mpval = row[["minPts"]]
    #   cvec = cluster_module(data.dist, algo="OPTICS", xi=xival, minPts=mpval)
    #   if(is.null(cvec)) cvec = rep(0, nrow(data.dist))
    #   names(cvec) = paste0("sample",1:length(cvec))
    #   c(param.algo="OPTICS", param.xi=xival, param.minPts=mpval,
    #     evaluation_module(data.dist, cvec),
    #     NMI=aricode::NMI(unlist(cvec), labels),
    #     ARI=mclust::adjustedRandIndex(unlist(cvec), labels),
    #     cvec)
    # })
    # df2 = bind_rows(df2)
  
    
    df3 = bind_rows(lapply(k, function(kval) {
      bind_rows(lapply(algos, function(algo) {
        bind_rows(lapply(agg, function(aggval) {
         if (algo == "hierarchical" | algo == "agnes") {
           cvec = cluster_module(data.dist, algo=algo, k=kval, agg_method=aggval)
           names(cvec) = paste0("sample",1:length(cvec))
           c(param.algo=algo, param.k=kval, param.agg_method=aggval, 
             evaluation_module(data.dist, cvec), 
             NMI=aricode::NMI(unlist(cvec), labels),
             ARI=mclust::adjustedRandIndex(unlist(cvec), labels),
             cvec)
         }
        }))
      }))
    }))
    
    # eval <- bind_rows(df1, df2, df3) %>% relocate(starts_with('param'))
    eval <- bind_rows(df1, df3) %>% relocate(starts_with('param'))
    # cvec_mat <- do.call(rbind, cvec_val)
    # out = list(eval.res=eval, cvec_val.res=cvec_val)
    # class(out) = 'run_module'
    return(eval)
}
```

# Running on Derma dataset

```{r derma_dataset, warning=FALSE, cache=TRUE}
data("derma")
dim(derma)
levels(as.factor(derma$ES))
head(derma)
```

Convert all Clinical and Histopathological variables to factors:

```{r}
derma.data <- derma %>% mutate(across(!ES, as.numeric)) %>% mutate_at(c("ES"), as.factor)
str(derma.data)
```


## Display the evaluation indices

```{r execute_run_all, warning=FALSE, cache=TRUE}
run_output <- run_all(data=derma.data[1:34], labels=derma.data$ES)
eval_res <-  dplyr::select(run_output, !starts_with("sample"))
reactable(eval_res,
          columns = list(cohesiveness = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         distinctiveness = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         avg_silwidth = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         max_diameter = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         min_separation = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         pearson_gamma = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         dunn_index = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         dunn2_index = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         entropy = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         wb_ratio = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         ch_index = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         widest_gap = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         sep_index = colDef(format = colFormat(locales = "en-US", digits = 2)),
                         index_C = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         index_G2 = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         index_G3 = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         index_S = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         NMI = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         ARI = colDef(format = colFormat(locales = "en-US", digits = 4))))

```

## Creates a matrix of the cluster assignments

```{r cvec_matrix, warning=FALSE, cache=TRUE}
cvec_mat <- as.matrix(dplyr::select(run_output, starts_with("sample")))
# cvec_mat[1,]
dim(cvec_mat)
```


# Correlation between evaluation indices

## Compare fpc avg silhouette width with Index_S (clusterSim) using pearson correlation

```{r silhouette_corr, warning=FALSE, cache=TRUE}
cor(eval_res$avg_silwidth, eval_res$index_S, method = "pearson", use = "complete.obs")
```

## Create a spearman's rank correlation pairwise heatmap to compare all the evaluation metrics

```{r heatmap, warning=FALSE, cache=TRUE}
corr_mat <- round(cor(eval_res[,7:ncol(eval_res)], method = "spearman", use = "pairwise.complete.obs"),2)
# print(corr_mat)

melted_corr_mat <- melt(corr_mat)

pheatmap(corr_mat, 
         display_numbers = TRUE,  # Display correlation values inside cells
         color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
         main = "Spearman Correlation Heatmap for Evaluation Metrics")

```

## Display top indices correlated to NMI

```{r nmi_corr, warning=FALSE, cache=TRUE}
melted_corr_mat %>% 
  filter(Var1 == "NMI" & !Var2 %in% c("NMI","ARI")) %>% 
  slice_max(value, n=5)
```


# Compare cluster assignments to the true classes of Derma dataset 

## Display the NMI and ARI indices of the first CVEC with true Derma labels

```{r true_indices, warning=FALSE, cache=TRUE}
aricode::NMI(cvec_mat[1,], derma.data$ES)
mclust::adjustedRandIndex(cvec_mat[1,], derma.data$ES)
```

## Display the NMI values for runs with k=6

```{r k_3, warning=FALSE, cache=TRUE}
eval_res_k_6 <- eval_res %>% 
  dplyr::filter(param.k == 6) %>%
  dplyr::select(starts_with('param') | c(NMI, ARI))

reactable(eval_res_k_6,
          columns = list(NMI = colDef(format = colFormat(locales = "en-US", digits = 4)),
                         ARI = colDef(format = colFormat(locales = "en-US", digits = 4))))
```


## NMI for Partitioning Clustering methods for different values of k

```{r partitioning_nmi, warning=FALSE, cache=TRUE}
data1 <- filter(eval_res, !param.algo %in% c("OPTICS","hierarchical","agnes"))
ggplot(data1, aes(x = param.k, y=NMI, fill = param.algo)) + 
  geom_col() + 
  facet_grid(vars(param.algo))
```

## NMI for Hierarchical Clustering methods for different values of k and aggregation method

```{r hierarchical_nmi, warning=FALSE, cache=TRUE}
data2 <- filter(eval_res, param.algo %in% c("hierarchical","agnes"))
ggplot(data2, aes(x = param.k, y=NMI, fill = param.algo)) + 
  geom_col() + 
  facet_grid(param.algo ~ param.agg_method)
```


## ARI for Partitioning Clustering methods for different values of k

```{r partitioning_ari, warning=FALSE, cache=TRUE}
data3 <- filter(eval_res, !param.algo %in% c("OPTICS","hierarchical","agnes"))
ggplot(data3, aes(x = param.k, y=ARI, fill = param.algo)) + 
  geom_col() + 
  facet_grid(vars(param.algo))
```

## ARI for Hierarchical Clustering methods for different values of k

```{r hierarchical_ari, warning=FALSE, cache=TRUE}
data4 <- filter(eval_res, param.algo %in% c("hierarchical","agnes"))
ggplot(data4, aes(x = param.k, y=ARI, fill = param.algo)) + 
  geom_col() + 
  facet_grid(param.algo ~ param.agg_method)
```



# Notes
- Append the NMI for all to the run all function. 
- Histogram for all of NMI. Split by algorithm. If it doesn't work for optics, create histogram for other algos. 
- Run both NMI to ARI.

